\texttt{\emph{Incertae sedis}}\texttt{ taxa in supertree construction}

### 2018-02-21 doc/incertae_sedis_latex2.tex
Export to TEX.

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{babel}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage[unicode=true]
 {hyperref}

@@

@@

\title{\texttt{Handling }\texttt{\emph{incertae sedis}}\texttt{ taxa in
propinquity}}
\maketitle

@@

Currently \emph{incertae sedis} taxa and their descendants are excluded
from the OpenTree synthetic tree. This is problematic, because the
OpenTree project seeks to build a comprehensive tree covering all
of life. However, including unplaced taxa within the current synthesis
framework would be equally problematic, since placement of taxa within
groups is currently considered to be a conflict with those groups.
This conflict causes both a loss of the taxon name for the conflicting
taxon, and a loss of structure in the synthesis tree.

@@

\subsection{Handling incertae sedis taxa}

To solve this problem, we describe an extension of our previous supertree
synthesis procedure that can use phylogeny information to place \emph{incertae
sedis} taxa into containing taxa without considering the containing
taxa to be broken. This extension requires that we first describe
an extended semantics for taxa in the presences of \emph{incertae
sedis}, which is not a trivial problem. We describe both an updated
semantics for conflict with taxonomy edges, and an updated semantics
for assigning taxonomy names to nodes in the synthetic tree.

\subsection{Incertae sedis taxa in taxonomies}

@@

\textbf{Taxonomy merging.} Not all \emph{incertae sedis} taxa in our
taxonomy are directly labeled \emph{incertae sedis} by taxonomists.
\emph{Incertae sedis} taxa can also result from automatic merging
of taxonomies to create the OpenTree taxonomy. For example, in Figure
3 of \citet{rees2017automated}, cases \#4 and \#6 illustrate examples
where merging of two taxonomies leads to a taxonomy with a taxon of
uncertain placement. The reason is primarily that if taxonomy $\taxonomy_{1}$
contains more levels of hierarchy than taxonomy $\taxonomy_{2}$,
then we must add internal nodes to $\taxonomy_{2}$ to align it to
$\taxonomy_{1}$. However, if taxonomy $\taxonomy_{2}$ contains more
leaves than $\taxonomy_{1}$, then it is unclear if these extra leaves
should be nested inside the additional internal nodes, or not. Thus,
the extra leaves are marked \emph{incertae sedis}.

\subsection{Formalism}

We consider a ranked collection $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of input trees and a single taxonomy tree $\taxonomy$. We have a
set of labels $\mathcal{L}$ that correspond to taxon names. Therefore
taxonomy node has a label, and every label correspond to a unique
taxonomy node.

@@

Incertae sedis taxa affect the synthesis procedure in two ways: they
affect what it means for a taxonomy edge to conflict with the synthesis
tree, and they affect what it means to place a taxon name on a particular
node in the synthesis tree. We examine each of these effects in turn.

@@

@@

@@

\section{Results}

Should we do this? We could say: 
\begin{itemize}
\item we placed $x$ incertae sedis taxa.
\item we avoiding breaking $y$ taxa that had IS taxa placed inside them.
\item we allowed $z_{1}$ new taxa into the synthesis tree that were incertae
sedis.
\item we allowed $z_{2}$ new taxa into the synthesis tree that are marked
as extinct.
\item some nodes have as many as $w$\emph{ incertae sedis} children, making
them unbrowseable when incertae sedis children are not excluded.
\item $v_{1}$ input trees w were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input trees w were previously excluded \emph{partially} because
they are nested within in an incertae sedis taxon.
\end{itemize}
Currently the numbers $z_{1}$ and $z_{2}$.

>>>>

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{babel}
\usepackage{verbatim}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage[unicode=true]
 {hyperref}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newenvironment{lyxcode}
{\par\begin{list}{}{
\setlength{\rightmargin}{\leftmargin}
\setlength{\listparindent}{0pt}% needed for AMS classes
\raggedright
\setlength{\itemsep}{0pt}
\setlength{\parsep}{0pt}
\normalfont\ttfamily}%
 \item[]}
{\end{list}}

@@

\title{\texttt{\emph{Incertae sedis}}\texttt{ taxa in supertree construction}}
\maketitle

@@

The OpenTree project seeks to build a comprehensive supertree tree
covering all of life. The approach is to combine information in published
phylogenies with a comprehensive taxonomy that supplies taxon names.
We describe an extension to our supertree method so that we can resolve
taxonomic uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa. We describe an updated semantics for conflict between
phylogeny edges and taxonomy edges and an updated semantics for assigning
taxonomy names to nodes in the synthetic tree.

\emph{TODO: add cases}

\paragraph{Taxonomy merging}

Not all \emph{incertae sedis} taxa in our taxonomy are directly labeled
\emph{incertae sedis} by taxonomists. \emph{Incertae sedis} taxa can
also result from automatic merging of taxonomies to create the OpenTree
taxonomy. For example, in Figure 3 of \citet{rees2017automated},
cases \#4 and \#6 illustrate examples where merging of two taxonomies
leads to a taxonomy with a taxon of uncertain placement. The reason
is primarily that if taxonomy $\taxonomy_{1}$ contains more levels
of hierarchy than taxonomy $\taxonomy_{2}$, then we must add internal
nodes to $\taxonomy_{2}$ to align it to $\taxonomy_{1}$. However,
if taxonomy $\taxonomy_{2}$ contains more leaves than $\taxonomy_{1}$,
then it is unclear if these extra leaves should be nested inside the
additional internal nodes, or not. Thus, the extra leaves are marked
\emph{incertae sedis}.

\paragraph{Formalism}

Formalism. We consider a ranked collection $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of input trees and a single taxonomy tree $\taxonomy$. We have a
set of labels $\mathcal{L}$ that correspond to taxon names. Therefore
taxonomy node has a label, and every label correspond to a unique
taxonomy node.

\section{Details \& Methods}

@@

\subsection{Taxonomy details}

@@

@@

In order to incorporate \emph{incertae sedis }taxa into a supertree
analyses, we first tackle the issue of the semantics of \emph{incertae
sedis} taxa. We discuss various different things the term can mean
within the literature, and settle on a specific usage here. We discuss
how to move from an operational ``you can move this guy here'' semantics
to a splits-based semantics. This affects both what it means for a
taxonomy edge to conflict with the synthesis tree, and what it means
to place a taxon name on a particular node in the synthesis tree.
Finally, after a supertree has been constructed, we consider how to
interpret the supertree as (perhaps) placing an \emph{incertae sedis}
taxon $A$ within its sister taxon $B$.

\subsection{Usage in the literature}

The term \emph{incertae sedis} can be used in a number of different
ways. For example,\emph{ }one author might describe \emph{incertae
sedis} genera that are asserted to be monophyletic, whereas another
author might describe \emph{incertae sedis} genera that may be interdigitated.

@@

\subsubsection{Comparison to operational definition}

An definition in terms of moving node $A$ to position $x$ could
cause trouble when moving multiple nodes, since the order of the movement
would matter.

Also, the splits-based semantics might not completely recover the
edit-operation semantics. The corner cases are ... .

@@

After a synthesis tree is constructed, we attach taxon names for to
the synthesis tree for each non-conflicting taxon. These names attach
at the MRCA of the include group. Then, for each taxon $A$ on the
synth tree, we associate with the first taxon $B$ found in a rootward
walk on the synthesis tree. If $A$ is not a descendant of $B$ on
the taxonomy, then we say that $A$ is \emph{placed} within $B$ by
the synthesis tree. While this may occur if $A$ is \emph{incertae
sedis}, more complex scenarios are possible. We note that if method
does not handle \emph{incertae sedis} taxa, then this scenario could
not occur because any taxa $B$ that $A$ is placed within would be
considered incompatible with the synthesis tree and so their names
would not be applied.

In order to fully describe a placement of $A$ within $B$ we consider
the sequence of taxa encountered on a walk from $A$ or $B$ up to
their MRCA $M$ on the taxonomy tree. Let us denote the sequence of
taxa encountered from $A$ as $A_{1},\ldots,A_{n},M$ where $A_{1}=A$.
Let $B_{1},\ldots,B_{m},M$ be the sequence of taxa encountered from
$B$ where $B_{1}=B$. Then $A_{n}$ must be \emph{incertae sedis},
and $A_{2},\ldots,A_{n}$ must be incompatible with the synthesis
tree (i.e. broken taxa). We then have that $A$ is placed successively
within $B_{m},B_{m-1},\ldots,B_{2},B_{1}$, and $m$ is the depth
of the placement of $A$.

One question is whether this approach is able to handle nested incertae
sedis taxa.

@@

\section{Results}
\begin{lyxcode}
\begin{comment}
\begin{lyxcode}
?ott6145835~{[}6145835{]}~<-~ott698424~{[}698424{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}

?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~

?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}

?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

?ott564110~{[}564110{]}~<-~ott939130~{[}939130{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~

?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~

?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~

?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~ott447802~{[}447802{]}~<-~ott23373~{[}23373{]}

?ott441413~{[}441413{]}~<-~ott1014022~{[}1014022{]}~placed~under~ott571842~{[}571842{]}~<-~ott997381~{[}997381{]}~<-~ott1014022~{[}1014022{]}~

?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~

?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~

?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
\end{lyxcode}
\end{comment}
\end{lyxcode}
Should we do this? We could say: 
\begin{itemize}
\item we placed 14 incertae sedis taxa.
\item we avoiding breaking 8 taxa that had IS taxa placed inside them.
\item we allowed $z_{1}$ new taxa into the synthesis tree that were incertae
sedis.
\item we allowed $z_{2}$ new taxa into the synthesis tree that are marked
as extinct.
\item some nodes have as many as $w$\emph{ incertae sedis} children, making
them unbrowseable when incertae sedis children are not excluded.
\item $v_{1}$ input trees were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input trees were previously excluded \emph{partially} because
they are nested within in an incertae sedis taxon.
\end{itemize}
Currently the numbers $z_{1}$ and $z_{2}$.
